### 2015-03-16 WhitePapers/revise_chunks/paper-body.tex
Minor updates to paper
Former-commit-id: 79ccdcd6271a12b6751ead4c5386947273d72eee

This paper is organized as follows. 
Section~\ref{sec:architecture} presents the 
current indexing method used within the library for chunking as well as 
modifications that were made to improve performance when a dataset is 
being extended. Section~\ref{sec:evaluation} presents performance evaluation 
results.
In Section~\ref{sec:related_work} we
first discuss related work before presenting conclusions and 
future research directions in Section~\ref{sec:conclusion}.

@@

B-trees are generally a good solution that is used for indexing data blocks
and one of the main components of file systems~\cite{Comer1979}~\cite{Folk1992}. 
Insertion of a new entry and lookup is realized in $O(\log_b{n})$, where $n$ is the
number of nodes and $b$ the order of the tree.
In the case of HDF5 datasets and in the original B-tree structure, referred to
as \textit{B-Tree version 1} (BT1), the record for locating the chunk stores the
following information:
\begin{itemize}
\item The coordinates of the chunk in the dataset's dataspace. This is used
as the key for locating chunks in the B-tree.
\item The size of the chunk, in bytes.
\item The address of the chunk in the file.
\item Additional metadata.
\end{itemize}

%One of the first issues is the redundancy of chunk size,
%inserted in every record, as the chunks are fixed size.
As shown in figure~\ref{fig:btree1}, each leaf points to the address of a
chunk. When a dataset is extended, a new entry is added, which may
imply a node split if the node where the record needs to be inserted is full. 
It is worth noting that most chunked datasets are either fixed dimension or
extend in only one dimension. Very few users take advantage of the ability to 
extend multiple dimensions of a dataset.
Because HDF5 datasets only allow their dimensions to be increased or 
decreased at their upper bound, a B-tree used to index chunks for a 1-D 
dataset will only ever insert or remove records from its right-most node.
One of the problems is that inserting/removing only from the right-most
node lets most records half empty, as illustrated in 
figure~\ref{fig:btree1}.

@@

For this particular case, indexing chunks requires a more appropriate data
structure, referred to as \textit{extensible array} (EA), and shown in
figure~\ref{fig:ea}. This structure can be seen as a dynamic array structure.
Brodnik et al's paper~\cite{Brodnik1999} addresses this exact problem,
showing how to implement a dynamic array with $O(1)$ append, shrink and lookup
operations, along with optimal metadata overhead for indexing the array elements.
For indexing chunks in HDF5 datasets, the array elements in the data blocks are
pointers to the chunks on disk.
As mentioned in the paper, the resizable array \textit{super blocks} are not actually 
constructs in the data structure, they are just shown to conceptually group the data 
blocks that the index block points to. The super blocks help show the pattern of 
changes to the data blocks: first the size of the data block doubles, then the 
number of data blocks of that size doubles, then the size doubles again, then the 
number of data blocks of that size doubles, etc.

A couple of things should be noted about the data structure in
figure~\ref{fig:ea}:
\begin{enumerate}
\item The super blocks for the first two data blocks are omitted and the pointers
in the index block point directly to the data blocks.
\item Constant time (i.e., $O(1)$) lookup, append and shrink are still in effect.
The number of I/O accesses to find any chunk address is always $2$ or $3$:
index block $\rightarrow$ super block $\rightarrow$ data block. In
actual operation, the index block will certainly be \textit{hot} enough to stay
in the HDF5 metadata cache, along with many/most of the super blocks,
making the average number of I/O accesses to retrieve a chunk address between
$1$ and $2$, in all likelihood.
\item Super blocks/data blocks which are not yet needed are not allocated and
have a \textit{nil} pointer in the index block/super block (respectively) that refers to them.
\item The height of the index block is constant and its height can be computed
at creation so that resizing it is unnecessary. This can be done by computing
the maximum number of chunks possible for the file's address range
(typically 64-bits) and deriving the maximum number of data/index blocks needed
to index that many blocks. For example, if there are $16$ chunks in the initial
data block, a chunk size of $2048$ (with a single unlimited dimension), and a
4-byte (integer) data element size, maximum height of the index block needed to
store $1.84\times 10^{19}$ bytes of elements (which is the maximum offset possible in a
64-bit file) is $47$.
\end{enumerate}

@@

With these data structures in place, when the chunked dataset has only one
unlimited maximum dimension, the extensible array indexing method is used,
allowing chunk access in $O(1)$ time. When it has more than one unlimited
maximum dimension sizes, the B-tree version 2 indexing method is used, allowing
chunk access in $O\log_b{n}$, with $b$ the order of the tree and $n$ the
number of nodes.

>>>>

This paper is organized as follows. 
Section~\ref{sec:architecture} presents the 
current indexing method used within the library for chunking as well as 
modifications that were made to improve performance when a dataset is 
being extended. Section~\ref{sec:evaluation} presents performance evaluation 
results. In Section~\ref{sec:related_work} we discuss related work before
presenting conclusions and future research directions in Section~\ref{sec:conclusion}.

@@

B-trees are generally used for indexing data blocks
and one of the main components of file systems~\cite{Comer1979}~\cite{Folk1992}. 
Insertion of a new entry and lookup is realized in $O(\log_b{n})$, where $n$ is the
number of nodes and $b$ the order of the tree.
In the case of HDF5 datasets and in the original B-tree structure, referred to
as \textit{B-Tree version 1} (BT1), the record for locating the chunk stores the
following information:
\begin{itemize}
\item The coordinates of the chunk in the dataset's dataspace. This is used
as the key for locating chunks in the B-tree.
\item The size of the chunk, in bytes.
\item The address of the chunk in the file.
\item Additional metadata.
\end{itemize}

%One of the first issues is the redundancy of chunk size,
%inserted in every record, as the chunks are fixed size.
As shown in figure~\ref{fig:btree1}, each leaf points to the address of a
chunk. When a dataset is extended, a new entry is added, which may
generate a node split if the node where the record needs to be inserted is full. 
It is worth noting that most chunked datasets are either fixed dimension or
extend in only one dimension. Very few users take advantage of the ability to 
extend multiple dimensions of a dataset.
Because HDF5 datasets only allow their dimensions to be increased or 
decreased at their upper bound, a B-tree used to index chunks for a 1-D 
dataset will only ever insert or remove records from its right-most node.
One of the problems is that inserting/removing only from the right-most
node lets most records half empty, as illustrated in 
figure~\ref{fig:btree1}.

@@

For this particular case, indexing chunks requires a more appropriate data
structure, referred to as \textit{extensible array} (EA), and shown in
figure~\ref{fig:ea}. This structure can be seen as a dynamic array structure.
Brodnik et al's paper~\cite{Brodnik1999} addresses this exact problem,
showing how to implement a dynamic array with $O(1)$ append, shrink and lookup
operations, along with optimal metadata overhead for indexing the array elements.
For indexing chunks in HDF5 datasets, the array elements in the data blocks are
pointers to the chunks on disk.
As mentioned in~\cite{Brodnik1999}, the resizable array \textit{super blocks} are not actually 
constructs in the data structure, they are just shown to conceptually group the data 
blocks that the index block points to. The super blocks help show the pattern of 
changes to the data blocks: first the size of the data block doubles, then the 
number of data blocks of that size doubles, then the size doubles again, then the 
number of data blocks of that size doubles, etc.

A couple of things should be noted about the data structure in figure~\ref{fig:ea}:
\begin{enumerate}
\item The super blocks for the first two data blocks are omitted and the pointers
in the index block point directly to the data blocks.
\item Constant time (i.e., $O(1)$) lookup, append and shrink are still in effect.
The number of I/O accesses to find any chunk address is always $2$ or $3$:
index block $\rightarrow$ super block $\rightarrow$ data block. In
actual operation, the index block will certainly be \textit{hot} enough to stay
in the HDF5 metadata cache, along with many/most of the super blocks,
making the average number of I/O accesses to retrieve a chunk address between
$1$ and $2$, in all likelihood.
\item Super blocks/data blocks, which are not yet needed, are not allocated and
have a \textit{nil} pointer in the index block/super block (respectively) that
refers to them.
\item The height of the index block is constant and its height can be computed
at creation so that resizing it is unnecessary. This can be done by computing
the maximum number of chunks possible for the file's address range
(typically 64-bits) and deriving the maximum number of data/index blocks needed
to index that many blocks. For example, if there are $16$ chunks in the initial
data block, a chunk size of $2048$ (with a single unlimited dimension), and a
4-byte (integer) data element size, the maximum height of the index block needed to
store $1.84\times 10^{19}$ bytes of elements (which is the maximum offset
possible in a 64-bit file) is $47$.
\end{enumerate}

@@

With these data structures in place, when the chunked dataset has only one
unlimited maximum dimension, the extensible array indexing method is used,
allowing chunk access in $O(1)$ time. When it has more than one unlimited
maximum dimension sizes, the B-tree version 2 indexing method is used, allowing
chunk access in $O(\log_b{n})$, with $b$ the order of the tree and $n$ the
number of nodes.
